---
title: "Assignment 09: Data Scraping"
author: "Sashoy Milton"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1

getwd() #Check working directory

#Load packages

library(tidyverse)
library(rvest)
library(ggplot2)
library(lubridate)
library(dplyr)

# Set theme

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2

durham_muncipal <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021')
durham_muncipal

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}

#3 Scaped and assigned values

water.system.name <- durham_muncipal %>% html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
  
pswid <- durham_muncipal %>% html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
  
ownership <- durham_muncipal %>% html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
  
max.withdrawals.mgd <- durham_muncipal %>% html_nodes('th~ td+ td') %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4 Create data-frame

local_water_supply.2021 <- data.frame("Month" = c("Jan", "May", "Sept", "Feb", "Jun",
                                  "Oct", "Mar", "Jul", "Nov", "Apr", "Aug","Dec"),
                             "Year" = rep(2021,12),
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>% 
  mutate(Water.System.Name = !!water.system.name,
         PSWID = !!pswid,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))


local_water_supply.2021 #Check dataset

#Check variable type

class(local_water_supply.2021$Date)

class(local_water_supply_reorder$Max_Withdrawals_mgd)


#5. Create line plot of maximum daily withdrawals

ggplot(local_water_supply.2021, aes(x = Date, y = max.withdrawals.mgd, group = 1)) +
  geom_line() +
  ylab ("Maximum Daily Withdrawals/ mgd") +
  xlab ("Months of the Year") +
  labs(title ="Max Daily Water Withdrawals 2021 in Durham")


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.


scrape_website <- function(the_pswid,the_year){
  
  
#Retrieve the website contents 
    
the_website <-read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',the_pswid,'&year=',the_year))
                               
#Set the element address variables 
  water_system_name_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  pswid_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  max_withdrawal_tag <- 'th~ td+ td'
  
#Scrape the data items
water.system.name <- the_website %>% html_nodes(water_system_name_tag) %>% html_text()
pswid <-the_website %>% html_nodes(pswid_tag) %>% html_text()
ownership <- the_website %>% html_nodes(ownership_tag) %>% html_text()
max.withdrawals.mgd <- the_website %>% html_nodes(max_withdrawal_tag) %>% html_text()
  
#Convert to a dataframe
 water_supply <- data.frame("Month" = c("Jan", "May", "Sept", "Feb", "Jun",
                                  "Oct", "Mar", "Jul", "Nov", "Apr", "Aug","Dec"),
                             "Year" = rep(2015,12),
                             "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>% 
 mutate(Water.System.Name = !!water.system.name,
         PSWID = !!pswid,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))
 
 Sys.sleep(1)

  #Return the dataframe
  return(water_supply)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7

# Extract Max Daily Withdrawals for Durham

local_water_supply.2015 <- scrape_website('03-32-010',2015)
local_water_supply.2015 # View data set

# Line plot of maximum daily withdrawals 

ggplot(local_water_supply.2015, aes(x = Date, y = Max_Withdrawals_mgd, group = 1)) +
  geom_line() +
  ylab ("Maximum Daily Withdrawals/ mgd") +
  xlab ("Month of the Year") +
  labs(title = ("Max Daily Water Withdrawals 2015 in Durham"))

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
##Extract Asheville data

local_water_supply_2015_Asheville <- scrape_website('01-11-010',2015)
local_water_supply_2015_Asheville #View data set

colnames(local_water_supply_2015_Asheville)
#Create a plot that compares Asheville's to Durham's water withdrawals

ggplot(local_water_supply_2015_Asheville, aes(x=Date, y=Max_Withdrawals_mgd, group = 1, color = Water.System.Name))+
  geom_line() +
  geom_line(data = local_water_supply.2015, aes(x = Date, y = Max_Withdrawals_mgd, group = 1, color = water.system.name)) +
    ylab ("Maximum Daily Withdrawals/ mgd") +
  xlab ("Month of the Year") +
  labs(title = ("Max Daily Water Withdrawals 2015 in Durham and Asheville"),
       color = "Water System Name")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}

#9

#Subset the facilities
#the_facility_ids_subset <-sample(the_facility_ids,20)

#Create a list of the year we want, the same length as the vector above
#the_years <- rep.int(2020,length(the_facility_ids_subset))

###########

#Set the inputs to scrape years 2015 to 2020 for the Asheville site "01-11-010"
the_years = rep(2010:2019)

#Subset the facilities
#the_years_subset <- sample(the_years,10)

my_facility = rep('01-11-010',10)

#Use lapply to apply the scrape function
# the_dfs <- lapply(X = the_years,
#                 FUN = scrape_website,
#                the_facility= my_facility)

#OR, use purrr's map function
the_dfs <- map2(my_facility,the_years,scrape_website)

#Conflate the returned dataframes into a single dataframe
the_df <- bind_rows(the_dfs)

#Plot, because it's fun and rewarding
ggplot(the_df,aes(x=Date,y=Avg_Withdrawals_mgd)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2020 Water usage data for",pswid),
       subtitle = water.system.name,
       y="Withdrawal (mgd)",
       x="Date")

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?