---
title: 'Assignment 5: Data Visualization'
author: "Sashoy Milton"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}


#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` 
with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a 
single PDF file.

The completed exercise is due on Friday, Oct 21st @ 5:00pm.



## Set up your session 

1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages.  

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
##1. Verify working directory

getwd()
setwd("C:/Users/sasho/Desktop/Environ Data Analytics/Env872 Workspace/EDA-Fall2022_SM")
#b.Load packages

library(tidyverse)
library(lubridate)
library (cowplot)

#c. Load datasets

lake_chemistry <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

litter_dat <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", 
                       stringsAsFactors = TRUE)

#2. Check that date is in date format

### Lake Chemistry data set

class(lake_chemistry$sampledate)
lake_chemistry$sampledate <- ymd(lake_chemistry$sampledate)
class(lake_chemistry$sampledate) # Recheck class of data
View (lake_chemistry)  #View data to view variable

### Litter data set

class(litter_dat$collectDate)
litter_dat$collectDate <- ymd(litter_dat$collectDate)
class(litter_dat$collectDate) #Recheck class of data
View (litter_dat)

```
## Define your theme

3. Build a theme and set it as your default theme.

```{r}

#3. Building a theme

my_theme <- theme_bw(base_size = 12) + 
  theme(axis.text = element_text(color = "black"), 
      legend.position = "top", legend.justification = "center") +
  theme(plot.title = element_text(hjust = 0.5))
theme_set(my_theme)
  
```

## Create graphs

For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r,warning=FALSE}

#4. Plot of total phosphorus by phosphate

  ggplot(lake_chemistry, aes(x = tp_ug, y = po4, color=lakename)) +
  geom_point() +
  geom_smooth (method = lm, color = "black") +
  xlab("Phosphate") +
  ylab ("Total phosphorus/ug") +
  ylim(0,200) + 
  ggtitle("Total phosphorus by phosphate for Peter and Paul Lakes") +
  labs (color = "Lake's Name")

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5. Boxplots of Temperature, TP, and TN

#### Separate boxplots



temp.plot <- ggplot(lake_chemistry,
                    aes(x = factor(month,levels = c(1:12), labels = month.abb),
                        y= temperature_C)) +
  geom_boxplot (aes(color = lakename)) +
  xlab ("Month") +
  ylab ("Temperature/ degrees in Celsius") +
  ggtitle(" Boxplot of Temperature of Peter and Paul Lake") +
  labs (color = "Name of Lake") +
  scale_x_discrete(drop = FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  

temp.plot #View graph

nitrogen.plot <- ggplot(lake_chemistry,
                        aes (x = factor(month,levels = c(1:12), labels = month.abb),
                             y = tn_ug)) +
  geom_boxplot (aes(color = lakename)) +
  xlab ("Month") +
  ylab ("Total Nitrogen/ug") +
  ggtitle(" Boxplot of Total Nitrogen of Peter and Paul Lake") +
  labs (color = "Name of Lake") +
  scale_x_discrete(drop = FALSE) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  

nitrogen.plot #View graph

phosphorus.plot <- ggplot(lake_chemistry,
                          aes (x = factor(month,levels = c(1:12), labels = month.abb),
                               y = tp_ug)) +
  geom_boxplot (aes(color = lakename)) +
  xlab ("Month") +
  ylab ("Total phosphorous/ ug") +
  ggtitle(" Boxplot of Total Phosphorus of Peter and Paul Lake") +
  labs (color = "Name of Lake") +
  scale_x_discrete(drop = FALSE) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
  

phosphorus.plot #View plot


```

```{r}

### Joint Plot

c.legend <- get_legend(temp.plot)

title <- ggdraw() + 
draw_figure_label("Boxplots of Temperature,Total Phosphorus 
                  and Nitrogen of Peter and Paul Lake",
                    position = "top.left")


combined_plot <- plot_grid(temp.plot + theme(legend.position = "none") + ggtitle(""), 
  nitrogen.plot + theme(legend.position = "none") + ggtitle(""), 
  phosphorus.plot + theme(legend.position = "none") + ggtitle(""), nrow = 1, align = "h")


plot_grid(combined_plot,c.legend, title, nrow = 3 , rel_heights = c(5, 1))

  
```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer:
    I observe that as we move from spring to summer (August), there is a general increase in median temperature levels for both Peter and Paul Lake, with Paul Lake experiencing higher overall temperatures. Beginning in fall (September), there is an overall decrease in median temperatures across both lakes, with Peter Lake experiencing lower temperatures except in the month of October.
    Generally, Paul Lake had lower total median nitrogen levels when compared to Peter Lake over all seasons measured (end of spring - early fall). Noticeably, the levels of total nitrogen for most of the data points were fairly consistent for both lakes over the same period.
    For total phosphorus, the median levels observed at Peter Lake were generally greater than the levels observed at Paul Lake over the observed seasons (late spring to early fall). Further, there was a steady increase in total median phosphorus levels observed at Peter Lake over the same period. Conversely, total median phosphorus levels at Paul Lake from late spring to early fall remained fairly steady.
    

6. [Niwot Ridge] Plot a subset of the litter data-set by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}

#6. 

subset_litter_dat <- litter_dat %>%
  filter(functionalGroup == "Needles") 

subset_litter_dat #View data

ggplot(subset_litter_dat,aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  xlab ("Month and Year of Collection") +
  ylab ("Dry Mass of Needle Litter") +
  ggtitle(" Distribution of Dry Mass of Needle Litter Over Time") +
  labs (color = "NLCD Class") +
  scale_x_date (limits = as.Date(c("2016-05-16","2019-09-25")), 
                 date_breaks = "2 months", date_labels = "%b %y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
 
#7

ggplot(subset_litter_dat, aes(collectDate, y = dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow = 3) +
  ylab ("Dry Mass of Needle Litter") +
  xlab ("Month and Year of Collection") +
  theme(strip.background = element_rect (color = "black", fill = "green")) +
   ggtitle(" Distribution of Dry Mass of Needle Litter Over Time") +
  scale_x_date (limits = as.Date(c("2016-05-16","2019-09-25")), 
                 date_breaks = "2 months", date_labels = "%b %y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: The plot with the NLCD classes separated into three facets rather than separated by color is more effective. This is as it is easier to decipher the differences in dry mass of needle litter for each of the NCLD classes over the observed period when they were separated into their own panel. When separated by color, it was difficult to interpret not only the variation observed at each time point for each NCLD class but to determine the differences in between each class.