---
title: "Assignment 2: Coding Basics"
author: "Sashoy Milton"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r sequence, mean and median}

#1. The function seq generates number sequences

num_seq <- seq(from = 1, to = 100)


#2. Find the mean of the sequence

mean_seq <- mean(num_seq)

#3. Find the median of the sequence

median_seq <- median(num_seq)

#4. Check if the mean is greater then the median

mean_seq>median_seq

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r vectors, matrix and functions}

student.names <- c("Peter", "Shantoy", "Ariel", "Larry") #Vector type: character
test_scores <- c(67, 85, 34, 92) #Vector type: numeric
pass_fail <-c(TRUE,TRUE,FALSE,TRUE) #Vector type: logical


#Create data frame
gradebook = data.frame(student.names,test_scores,pass_fail)
gradebook


#Assign column names
column_names <- c("Student names", "Test score/100", "Pass?")

names(gradebook) <- column_names
gradebook
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame unlike a matrix contains several vector types. As seen in the dataframe above which contains character, numeric and logical vectors. Conversely, a matrix can only store one type of data even though it is two dimensional.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}

check_pass <- function(vector) {
  if (vector>50) {
    print(TRUE)
  }
  else {
    print(FALSE)
  }
}

check_pass(test_scores)

####


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
