---
title: "Assignment 7: Time Series Analysis"
author: Sashoy Milton
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on time series analysis. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A07_TimeSeries.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 16 at 11:59 pm.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE, warning = FALSE}
#1.

## Load library

library(tidyverse)
library(lubridate)
library (zoo)
library (trend)
library(Kendall)


## Check working directory

getwd()
setwd("C:/Users/sasho/Desktop/Environ Data Analytics/Env872 Workspace/EDA-Fall2022_SM")

##Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)


## Load datasets

EPAair_O3_GaringerNC2010_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv")
summary(EPAair_O3_GaringerNC2010_raw)
unique(EPAair_O3_GaringerNC2010_raw)

EPAair_O3_GaringerNC2011_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv")

EPAair_O3_GaringerNC2012_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv")

EPAair_O3_GaringerNC2013_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv")

EPAair_O3_GaringerNC2014_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv")

EPAair_O3_GaringerNC2015_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv")

EPAair_O3_GaringerNC2016_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv")

EPAair_O3_GaringerNC2017_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv")

EPAair_O3_GaringerNC2018_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv")

EPAair_O3_GaringerNC2019_raw <- read_csv("Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv")


# Combine data into one dataset

GaringerOzone_list <- list(EPAair_O3_GaringerNC2010_raw,EPAair_O3_GaringerNC2011_raw,EPAair_O3_GaringerNC2012_raw,EPAair_O3_GaringerNC2013_raw,EPAair_O3_GaringerNC2014_raw,EPAair_O3_GaringerNC2015_raw,EPAair_O3_GaringerNC2016_raw,EPAair_O3_GaringerNC2017_raw, EPAair_O3_GaringerNC2018_raw, EPAair_O3_GaringerNC2019_raw)


GaringerOzone <- Reduce(function(x,y) merge (x, y, all = TRUE), GaringerOzone_list)
dim(GaringerOzone) # Check dimensions


```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3. Set date column as date class

GaringerOzone$Date <- mdy(GaringerOzone$Date)
class(GaringerOzone$Date) #Check class
summary(GaringerOzone)


# 4. Subset data

GaringerOzone_processed <- GaringerOzone %>%
  select(Date, `Daily Max 8-hour Ozone Concentration`, DAILY_AQI_VALUE)

View(GaringerOzone)

# 5. Explore data
Days <- as.data.frame(seq(ymd('2010-01-01'), ymd('2019-12-31'), by = 'days')) #Daily Dataset
head(Days) # Check data


## Get column name

colnames(Days)

## Rename column days to date

Days <- Days %>% 
  rename(Date = "seq(ymd(\"2010-01-01\"), ymd(\"2019-12-31\"), by = \"days\")")

head(Days) #Check column name change


# 6.Join days and GaringerOzone

GaringerOzone <- left_join(Days, GaringerOzone_processed) #Rewrite GaringerOzone
dim(GaringerOzone) # Check dimensions
head(GaringerOzone) #Check data
summary(GaringerOzone)

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7. Plot line graph

class(GaringerOzone$`Daily Max 8-hour Ozone Concentration`)

ggplot(GaringerOzone, aes(x = Date, y = `Daily Max 8-hour Ozone Concentration`))+
  geom_line() +
  ylab("Daily Max 8-hour Ozone Concentration (ppm)")+
   geom_smooth (method = "lm") +
  ggtitle ("Ozone Concentrations (ppm) Over A Period From 2010 to 2019")
  
```

>Answer: There is a slight decline in ozone concentration (ppm) seen over the period. 

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8. Linear interpolation


GaringerOzone_interpolate <- GaringerOzone %>%
  mutate(Ozone_Concentration_cont = na.approx(`Daily Max 8-hour Ozone Concentration`))

summary(GaringerOzone_interpolate) # View data

```

> Answer: We did not use the piecewise function as we wanted to fill the gaps by predicting the value that may be miss from the trend based what  we assummed may fall between two measurements rather than filling the data with data that is equal to the measurement neasrest to the date. This approach of "connecting the dots" rather than nearest nesighbor" is perhaps better suited because we see a slight decline in the trend of the data. The spline approach is not suited for this interpolation as the trend is more consistent with a striaght line rather than a quadratic function.


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r, warning=FALSE}
#9.

GaringerOzone.monthly <- GaringerOzone_interpolate %>%
  mutate(Year = year(Date)) %>%
  mutate(Month = month(Date)) %>%
  mutate (Month_Year = my(paste0(Month,"-", Year))) %>%
  group_by(Month,Year, Month_Year) %>%
  summarise(MeanOzone = mean(Ozone_Concentration_cont))


head(GaringerOzone.monthly) # View data

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10.

#Daily time series

GaringerOzone.daily.ts <- ts(GaringerOzone_interpolate$Ozone_Concentration_cont, start = c(2010,1), frequency = 365)
GaringerOzone.daily.ts

#Monthly time series

Garinger.monthly.ts <- ts(GaringerOzone.monthly$MeanOzone, start = c(2010,1), frequency = 12)
Garinger.monthly.ts

```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11.

#Decompose the daily time series

GaringerOzone.daily_decomposed <- stl(GaringerOzone.daily.ts, s.window = "periodic")
plot(GaringerOzone.daily_decomposed)

#Decompose the monthly time series

Garinger.monthly.decomposed <- stl(Garinger.monthly.ts, s.window = "periodic")
plot(Garinger.monthly.decomposed)

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12. Run the seasonal Mann-Kendall


GaringerOzone.monthly_trend <- Kendall::SeasonalMannKendall(Garinger.monthly.ts)
GaringerOzone.monthly_trend
summary(GaringerOzone.monthly_trend)

```

> Answer: In this case, the seasonal Mann-Kendall is the most appropriate as there is a seasonal component to the time series.

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}

# 13. Plot 

ggplot(GaringerOzone.monthly, aes(x = Month_Year, y = MeanOzone)) +
  geom_line() +
  geom_point() +
  xlab("Time") +
  ylab ("Mean Monthly Ozone Concentrations (ppm)") +
  ggtitle (" Mean Monthly Ozone Concentrations Over Time") +
  geom_smooth(method = lm)
  
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: In the original plot, there was a slight decline seen in ozone concentrations over time. The results of the Mann-Kendall
test indicates that the decline seen is insignificant, p =0.16323 and tau = -0.1, for the monthly trend.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15

Garinger.monthly.decomposed$time.series #View timeseries

deseason_Garinger <- Garinger.monthly.ts - Garinger.monthly.decomposed$time.series[,1]
plot(deseason_Garinger) # Check Seasonality

#16

GaringerOzone.monthly_trend.deseason <- MannKendall(deseason_Garinger)
summary(GaringerOzone.monthly_trend.deseason)


```

> Answer: My results remain synonymous with the output obtained with the seasonal Mann-Kendall. These results also indicate that the slight decline seen in Ozone concentrations over the period of 2010 - 2019 was not significant, p = 0.10388 and tau = 0.1.
