---
title: 'Assignment 4: Data Wrangling'
author: "Sashoy Milton"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Set up your session 

1. Check your working directory, load the `tidyverse` and `lubridate` packages, 
and upload all four raw data files associated with the EPA Air dataset, being sure 
to set string columns to be read in a factors. See the README file for the EPA air 
datasets for more information (especially if you have not worked with air quality 
data previously). 

```{r, load packages, message=FALSE}

#Load packages

library(tidyverse)
library(lubridate)
library (skimr)

#Load data

EPA.Air.NC18 <- read.csv("./EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)
EPA.Air.NC19 <- read.csv("./EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)
EPA.Air.PM25_NC18 <- read.csv("./EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)
EPA.Air.PM25_NC19 <- read.csv("./EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)

```

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}

#1. Air Data

############# Explore dimensions #####################

dim(EPA.Air.NC18)
dim (EPA.Air.NC19)

############# Explore column names and structure of databases ###########

summary(EPA.Air.NC18)
# skim(EPA.Air.NC18)
head(EPA.Air.NC18)
tail(EPA.Air.NC18)

#########################

summary(EPA.Air.NC19)
# skim(EPA.Air.NC19)
head(EPA.Air.NC19)
tail(EPA.Air.NC19)

#2. PM 2.5 Data

############# Explore dimensions #####################

dim(EPA.Air.PM25_NC18)
dim (EPA.Air.PM25_NC19)

############# Explore column names and structure of databases ###########

summary(EPA.Air.PM25_NC18)
# skim(EPA.Air.PM25_NC18)
head(EPA.Air.PM25_NC18)
tail(EPA.Air.PM25_NC18)


#########################

summary(EPA.Air.PM25_NC19)
# skim(EPA.Air.PM25_NC19)
head(EPA.Air.PM25_NC19)
tail(EPA.Air.PM25_NC19)

```

## Wrangle individual datasets to create processed files.
3. Change date to date

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, 
AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" 
(all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file
names as the raw files but replace "raw" with "processed".

```{r}
#3. Change date to date

########################### Air Data #########################

class(EPA.Air.NC18$Date)  #Check class of variable

EPA.Air.NC18$Date <- mdy(EPA.Air.NC18$Date)
class(EPA.Air.NC18$Date) #Check change of class
View(EPA.Air.NC18)

#########################

class(EPA.Air.NC19$Date) #Check class of variable

EPA.Air.NC19$Date <- mdy(EPA.Air.NC19$Date)
class(EPA.Air.NC19$Date) #Check change of class
View(EPA.Air.NC19)

########################## PM 2.5 ##########################

class(EPA.Air.PM25_NC18$Date) #Check class of variable

EPA.Air.PM25_NC18$Date <- mdy(EPA.Air.PM25_NC18$Date)
class(EPA.Air.PM25_NC18$Date) #Check change of class
View(EPA.Air.PM25_NC18)
#########################
  
class(EPA.Air.PM25_NC19$Date) #Check class of variable

EPA.Air.PM25_NC19$Date <- mdy(EPA.Air.PM25_NC19$Date)
class(EPA.Air.PM25_NC19$Date)  #Check change of class
View(EPA.Air.PM25_NC19)

#4. Select Columns

#Select the following columns: Date, DAILY_AQI_VALUE, Site.Name,
#AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

########################### Air Data #########################

processed_EPA.Air.NC18 <- EPA.Air.NC18[c("Date", "DAILY_AQI_VALUE", "Site.Name", 
            "AQS_PARAMETER_DESC", "COUNTY","SITE_LATITUDE","SITE_LONGITUDE" )]


##############

processed_EPA.Air.NC19 <- EPA.Air.NC19[c("Date", "DAILY_AQI_VALUE", "Site.Name",
          "AQS_PARAMETER_DESC", "COUNTY","SITE_LATITUDE","SITE_LONGITUDE" )]

######################### PM 2.5 #################################

processed_EPA.Air.PM25_NC18 <- EPA.Air.PM25_NC18 [c("Date", "DAILY_AQI_VALUE", 
"Site.Name", "AQS_PARAMETER_DESC", "COUNTY","SITE_LATITUDE","SITE_LONGITUDE" )]

#################

processed_EPA.Air.PM25_NC19 <- EPA.Air.PM25_NC19 [c("Date", "DAILY_AQI_VALUE", 
          "Site.Name", "AQS_PARAMETER_DESC", "COUNTY","SITE_LATITUDE",
          "SITE_LONGITUDE" )]

#5. Filling all cells in AQS_PARAMETER_DESC with "PM2.5"

processed_EPA.Air.PM25_NC18$AQS_PARAMETER_DESC <- "PM2.5"
  
processed_EPA.Air.PM25_NC19$AQS_PARAMETER_DESC <- "PM2.5"

#6. Save four processed data-sets in the Processed folder

write.csv(processed_EPA.Air.NC18, 
      file = "C:/Users/sasho/Desktop/Environ Data Analytics/Env872 Workspace/EDA-Fall2022_SM/Data/Processed/processed_EPA.Air.NC18.csv",
      row.names = FALSE)
          
write.csv(processed_EPA.Air.NC18,  
          file = "C:/Users/sasho/Desktop/Environ Data Analytics/Env872 Workspace/EDA-Fall2022_SM/Data/Processed/processed_EPA.Air.NC19.csv",
          row.names = FALSE)
          
write.csv(processed_EPA.Air.NC18,  
          file = "C:/Users/sasho/Desktop/Environ Data Analytics/Env872 Workspace/EDA-Fall2022_SM/Data/Processed/processed_EPA.Air.PM25_NC18.csv",
          row.names = FALSE)
          
write.csv(processed_EPA.Air.NC18,
          file = "C:/Users/sasho/Desktop/Environ Data Analytics/Env872 Workspace/EDA-Fall2022_SM/Data/Processed/processed_EPA.Air.PM25_NC19.csv",
          row.names = FALSE)


```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical
prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the 
following conditions: 

- Include all sites that the four data frames have in common: "Linville Falls", 
"Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School",
"Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne",
"Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` 
can figure out common factor levels)

- Some sites have multiple measurements per day. Use the split-apply-combine 
strategy to generate daily means: group by date, site, aqs parameter, and county. 
Take the mean of the AQI value, latitude, and longitude.
- Add columns for "Month" and "Year" by parsing your "Date" column 
(hint: `lubridate` package)
- Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate 
columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7. Check that column names are the same

colnames(processed_EPA.Air.NC18)
colnames(processed_EPA.Air.NC19)
colnames(processed_EPA.Air.PM25_NC18)
colnames(processed_EPA.Air.PM25_NC19)

#Combine with rbind

combined_EPA.data <- rbind(processed_EPA.Air.NC18, processed_EPA.Air.NC19,
processed_EPA.Air.PM25_NC18, processed_EPA.Air.PM25_NC19)

#8. Using the pipe function

filtered_combined_EPA.data <- combined_EPA.data %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", 
 "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain",
                   "West Johnston Co.", "Garinger High School", "Castle Hayne", 
                "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  group_by(AQS_PARAMETER_DESC,Date,Site.Name,COUNTY) %>%
  summarise(DAILY_AQI_VALUE = mean(DAILY_AQI_VALUE), 
            SITE_LATITUDE = mean(SITE_LATITUDE),
            SITE_LONGITUDE = mean(SITE_LONGITUDE)) %>%  
             mutate(Month = month(Date),
              Year = year(Date))


#9. Spreading Data sets

filtered_combined_EPA.data2 <- filtered_combined_EPA.data %>%
  pivot_wider(names_from = AQS_PARAMETER_DESC, values_from = DAILY_AQI_VALUE)

filtered_combined_EPA.data2  #Check spreading

#10. Call up the dimensions of tidy dataset

dim(filtered_combined_EPA.data2)

#11. Save your dataset

write.csv(filtered_combined_EPA.data2,  
          file = "C:/Users/sasho/Desktop/Environ Data Analytics/Env872 Workspace/EDA-Fall2022_SM/Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv",
          row.names = FALSE)

```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. 
Data should be grouped by site, month, and year. Generate the mean AQI values 
for ozone and PM2.5 for each group. Then, add a pipe to remove instances where 
a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a + b

filtered_combined_EPA.data3 <- filtered_combined_EPA.data2 %>%
  group_by(Site.Name, Month, Year) %>%
  summarize (Ozone = mean(Ozone),
             PM2.5 = mean (PM2.5)) %>%
  drop_na(Ozone) %>%
  drop_na(PM2.5)

filtered_combined_EPA.data3 # View data
  

#13.

dim (filtered_combined_EPA.data3)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: In this case, we wanted to remove all the NA values from the dataset 
(which also deletes the record within which the NA was found)
instead of keeping the NA values but not incorporating them into our calculations.
